---
title: "2019 CA Population"
author: "Savet Hong"
date: "`r Sys.Date()`"
output: html_document
---

```{r library, include=FALSE}
library(censusapi)
library(tidyverse)
library(choroplethr)
library(choroplethrMaps)

Sys.getenv("CENSUS_KEY")

```


```{r setup, include=FALSE}

# Get Metadata list of group variables B01001
group_B01001 <- listCensusMetadata(
  name = "acs/acs5",
  vintage = 2017,
  type = "variables",
  group = "B01001")
head(group_B01001)

# Create df of labels from metadata
agesexlab <- group_B01001 %>%
  separate(label, c("type", "total", "gender", "age_group"), "!!") %>%
  mutate(start_age = sub("^([0-9]+) .*", "\\1", age_group),
         start_age = replace( start_age,start_age == "Under 5 years", 0),
         end_age = sub("^([0-9]+) .+ ([0-9]+) .*", "\\2", age_group),
         end_age = sub("^([0-9]+) .*", "\\1", end_age),
         end_age = replace( end_age, end_age == "Under 5 years", 5),
         start_age = as.numeric(start_age),
         end_age = as.numeric(end_age),
         gender = replace(gender, is.na(gender) & concept == "SEX BY AGE", "Both")) %>%
  select(-(predicateType:predicateOnly)) %>%
  filter(grepl("^Estimate", type))
```


```{r data19, include=FALSE}
# Get 2019 CA County Population Estimates: Sex by Age
acs19pop <- getCensus(
  name = "acs/acs5", 
  vintage = 2019, 
  vars = c("NAME", "group(B01001)"), 
  region = "county:*", 
  regionin = "state:06"
)

# Categorize 3 age group for analysis
acs19pop_clean <- acs19pop %>%
  gather(name, values, contains("B01001")) %>%
  separate(NAME, c("county_name", "state_name"), ",") %>%
  inner_join(agesexlab, by = "name") %>%
  mutate(age_band = case_when( start_age < 18 ~ "Age 0-17",
                               start_age >= 18 & start_age < 65 ~ "Age 18-64",
                               start_age >= 65 ~ "Age 65+",
                               TRUE ~ "NA"))

# Get Total Population and add mega region
capop <- acs19pop_clean %>%
  filter(gender == "Both") %>%
  select(county, county_name, values) %>%
  rename(Population = values) %>%
  mutate(Population = as.numeric(Population),
         id = sub(" County$", "", county_name),
         north_south = case_when(id %in% c("Imperial", "Kern", "Los Angeles", "Orange",
                                           "Riverside", "San Bernardino", "San Diego", 
                                           "Santa Barbara", "San Luis Obispo", "Ventura")
                                 ~ "Southern California",
                                 TRUE ~ "Northern California"),
         megregion = case_when(id %in% c("Alameda", "Contra Costa", "Marin", 
                                         "Napa", "San Mateo", "Santa Clara", 
                                         "Solano", "Sonoma", "San Francisco") ~
                                 "Bay Area",
                               id %in% c("Sacramento", "Yolo", "El Dorado", "Placer",
                                         "Sutter", "Yuba", "Nevada") ~
                                 "Greater Sacramento",
                               id %in% c("Imperial", "Kern", "Los Angeles", "Orange",
                                           "Riverside", "San Bernardino", "San Diego", 
                                           "Santa Barbara", "San Luis Obispo", "Ventura")
                                 ~ "Southern California" ,
                               TRUE ~ "Other")) %>%
  select(-id)

# Calculate proportion
acs19group <- acs19pop_clean %>%
  mutate(values = as.numeric(values)) %>%
  filter(age_band != "NA") %>% # Remove population for both gender 
  group_by(state, county, county_name, state_name, GEO_ID, age_band) %>%
  summarise(pop = sum(values)) %>%
  group_by(state, county, county_name, state_name, GEO_ID) %>%
  mutate( prop_county = 100 * round(pop / sum(pop), 3)) %>% # Percent of population within county
  group_by(state, state_name, age_band) %>%
  mutate(prop_gp_state = 100 * round(pop / sum(pop), 3)) %>% # Percent of population within age_band in state
  group_by(state, state_name) %>%
  mutate(prop_state = 100 * round(pop / sum(pop), 3)) # Percent of state's population

## Get longitude and laditude for mapping counties
acs19mapdf <- acs19group %>%
  mutate(region = substr(GEO_ID, 11, 14),
         region = as.numeric(region),
         id = sub(" County$", "", county_name),
         id = tolower(id),
         value = prop_county)

#Combine everything together
ca_counties <- map_data("county", "california") %>% 
  select(lon = long, lat, group, id = subregion) %>%
  left_join(acs19mapdf, by = "id")


```

### Overview

In 2019, California total population was about `r paste(format(round(sum(capop$Population)/1e6, 1), trim = TRUE), "million")` people, and `r paste0(100 * round(sum(capop$Population[capop$north_south=="Southern California"])/sum(capop$Population), 3), "%")` of them resided in Southern California (SoCal). In particular, Los Angeles county accounts for `r paste0(100 * round(sum(capop$Population[capop$county=="037"])/sum(capop$Population), 3), "%")` of the state's population.


```{r, echo=FALSE, message=FALSE, fig.width=8, fig.height=9}
cntylab <- ca_counties %>%
  mutate(id = stringr::str_to_title(id)) %>%
  group_by(id, group) %>%
  summarise(lon = mean(lon), lat = mean(lat))
  
# ggplot(ca_counties, aes(lon, lat, group = group)) +
#   geom_polygon(fill = NA, colour = "grey60") +
#   geom_text(data = cntylab, aes(label = id) ) +
#   theme_void() +
#   theme(strip.background = element_blank()) 

popmap <- ca_counties %>%
  select(lon:GEO_ID) %>%
  distinct() %>%
  right_join(capop) %>%
  mutate(`Population (100,000)` = round(Population/1e5, 1))

p1 <- ggplot(popmap, aes(lon, lat)) +
  geom_polygon(aes(group = group, fill = north_south), colour = "grey60") +
  scale_fill_manual(values = c("lightblue","pink")) +
  geom_text(data = cntylab, aes(label = id) ) +
  theme_void() +
  theme(strip.background = element_blank()) 


p2 <- ggplot(popmap, aes(lon, lat, group = group)) +
  geom_polygon(aes(fill = `Population (100,000)`), colour = "grey60") +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  geom_text(data = cntylab, aes(label = id),  size = 1.5 ) +
  theme_void() +
  theme(strip.background = element_blank()) 

#cowplot::plot_grid(p1, p2, labels = "")

print(p2)
```

#### Age Distribution of State Population

```{r, include= FALSE}
agreg <- left_join(acs19group, capop, by = c("county", "county_name")) %>%
  #mutate(megregion = replace(megregion, county == "019", "Fresno")) %>%
  ungroup()


propreg <- agreg %>%
  filter(age_band == "Age 18-64") %>%
  group_by(megregion) %>%
  summarise(pop = sum(pop), Population = sum(Population)) %>%
  mutate(per_reg = pop / Population,
         perall = pop / sum(Population),
         perage = pop / sum(pop))

```

The population of Southern California is largely comprised of people aged 18-64 (`r 100 * round(propreg$per_reg[propreg$megregion == "Southern California"],3)`%). This region's working age adults (18-64) accounts for `r 100 * round(propreg$perage[propreg$megregion == "Southern California"],3)`% of all working age adults in California (`r 100 * round(propreg$perall[propreg$megregion == "Southern California"],3)`% of entire population). Other areas of the state that have high concentration of working age adults include the Bay Area (`r 100 * round(propreg$perage[propreg$megregion == "Bay Area"],3)`% of age group, `r 100 * round(propreg$perall[propreg$megregion == "Bay Area"],3)`% of state's population) and Greater Sacramento (`r 100 * round(propreg$perage[propreg$megregion == "Greater Sacramento"],3)`% of age group, `r 100 * round(propreg$perall[propreg$megregion == "Greater Sacramento"],3)`% of state's population).

```{r stateprop, echo=FALSE, message=FALSE, fig.width=10}
ggplot(ca_counties, aes(lon, lat, group = group)) +
  geom_polygon(aes(fill = prop_state))+
  scale_fill_gradientn(name = "Percentage", colours = rev(heat.colors(2))) +
  theme_void() +
  theme(strip.background = element_blank()) +
  facet_grid(. ~ age_band)
```


#### Age Distribution Within County

When considering each county's age population distribution, it is no surprise that each county's residents are largely working age adults (about 60%). What is interesting is that some counties have a higher share of children, mainly counties in the San Joaquin Valley. Whereas, counties on the northern and eastern parts of California have a higher share of elderly adults.


```{r cnty_prop, echo=FALSE, message=FALSE, fig.width=10}
ggplot(ca_counties, aes(lon, lat, group = group)) +
  geom_polygon(aes(fill = prop_county))+ # color = "white")+
  scale_fill_gradientn(name = "Percentage", colours = rev(heat.colors(2))) +
  #scale_fill_viridis_c(name = "Percentage", option = "A") +
  theme_void() +
  theme(strip.background = element_blank()) +
  facet_grid(. ~ age_band) + 
  theme()

```



```{r ageprop, echo=FALSE, message=FALSE, fig.width=10, include=FALSE}

## Distribution Within Age group

ggplot(ca_counties, aes(lon, lat, group = group)) +
  geom_polygon(aes(fill = prop_gp_state))+
  scale_fill_gradientn(name = "Percentage", colours = rev(heat.colors(2))) +
  #scale_fill_gradient2() +
  #scale_fill_gradientn(colours = wes_palette("Darjeeling2", 10, type = "continuous")) +
  #scale_fill_viridis_c() +
  theme_void() +
  theme(strip.background = element_blank()) +
  facet_grid(. ~ age_band)
```



